---
title: "PADOG Pathway Enrichment Analysis"
author: "Daniel Garbozo"
date: "2025-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)
```

# PADOG Pathway Enrichment Analysis Workflow

## Overview
This workflow performs Pathway Analysis by Differential Gene Expression Ranking (PADOG) on count matrix data with multiple gene set collections and pairwise comparisons.

**Method:** PADOG ranks genes by their differential expression and tests whether pathway genes are ranked significantly higher than expected by chance.

**Output:** CSV and RDS files with pathway enrichment results, including FDR-adjusted p-values and multiple comparison results.

**Author:** Daniel Garbozo  
**Date:** February 20, 2025

---

## Section 1: Required Libraries

Load all required R packages. If any are missing, uncomment the `install.packages()` or `BiocManager::install()` lines.

```{r setup-libraries, message=FALSE, warning=FALSE}
# Core analysis packages
library(PADOG)          # Pathway analysis
library(limma)          # For voom normalization

# Data handling packages
library(readxl)         # Read Excel files
library(dplyr)          # Data manipulation
library(stringr)        # String processing

# Parallel processing
library(parallel)       # Base parallel tools
library(doParallel)     # Foreach parallelization

# Installation (uncomment if needed):
# install.packages(c("readxl", "dplyr", "stringr", "doParallel"))
# BiocManager::install(c("PADOG", "limma"))
```

---

## Section 2: Configure Paths (CUSTOMIZE THIS SECTION)

Edit the paths below to match your system and data locations. These are the **only** lines that should need modification.

```{r config-paths}
# ================================================================
# CONFIGURATION SECTION: Update these paths for your data
# ================================================================

# Base directory containing all analysis data
main_dir <- "/path/to/your/analysis/directory"

# Directory containing GMT files (gene sets from MSigDB or similar)
# Expected: *.gmt files (except total_genesets.gmt which will be excluded)
genesets_dir <- paste0(main_dir, "/gene_sets/")

# Output directory where results will be saved
# Subdirectories will be created for each gene set collection
output_dir <- paste0(main_dir, "/PADOG_results/")

# Input expression data file (TSV format)
# Expected: Tab-separated file with genes as rows, samples as columns
#           First column should contain gene symbols (used as rownames)
#           Must include a 'symbol' column for gene names
expression_file <- paste0(main_dir, "/expression_data.tsv")

# Sample metadata file (Excel format)
# Expected columns: id, group, gp
# - id: Sample identifiers (must match column names in expression_file)
# - group: Broad group assignment
# - gp: Group prefix used in comparisons (e.g., "Ap", "Bp", "An", "Bn")
metadata_file <- paste0(main_dir, "/sample_metadata.xlsx")

# ================================================================
# Verify paths exist
# ================================================================

if (!dir.exists(genesets_dir)) {
  stop("Gene sets directory not found: ", genesets_dir)
}

if (!file.exists(expression_file)) {
  stop("Expression file not found: ", expression_file)
}

if (!file.exists(metadata_file)) {
  stop("Metadata file not found: ", metadata_file)
}

# Create output directory if it doesn't exist
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

cat("Configuration Summary:\n")
cat("  Gene sets directory: ", genesets_dir, "\n")
cat("  Expression file: ", expression_file, "\n")
cat("  Metadata file: ", metadata_file, "\n")
cat("  Output directory: ", output_dir, "\n\n")
```

---

## Section 3: Define Comparisons and Load Data

```{r prepare-data}
# ================================================================
# DEFINE COMPARISONS
# ================================================================
# Format: "GroupA_GroupB" where GroupA and GroupB match values in metadata$gp
# These will be used to subset samples and create design matrices

comparisons <- c("Bp_Ap", "Bp_An", "Ap_An", "Bp_Bn", "Bn_An")
cat("Comparisons to analyze:", paste(comparisons, collapse = ", "), "\n\n")

# ================================================================
# LOAD GENE SET FILES (GMT)
# ================================================================

# List all GMT files (excluding total_genesets.gmt)
genesets_files <- list.files(genesets_dir, pattern = "*.gmt", full.names = TRUE)
genesets_files <- genesets_files[!grepl("total_genesets.gmt", genesets_files)]

if (length(genesets_files) == 0) {
  stop("No GMT files found in: ", genesets_dir)
}

cat("Found", length(genesets_files), "gene set collections:\n")
for (f in genesets_files) {
  cat("  -", basename(f), "\n")
}
cat("\n")

# ================================================================
# LOAD SAMPLE METADATA
# ================================================================

metadata <- read_xlsx(metadata_file) %>%
  select(id, group, gp) %>%
  rename(Sample = id)

rownames(metadata) <- metadata$Sample

cat("Loaded", nrow(metadata), "samples\n")
cat("Sample groups found:", unique(metadata$gp), "\n\n")

# ================================================================
# LOAD EXPRESSION MATRIX
# ================================================================

# Read expression data (tab-separated, gene symbols as rownames)
PadMat_full <- read.csv(expression_file, sep = "\t", header = TRUE)

# Remove genes without gene symbols (missing values in 'symbol' column)
PadMat_full <- PadMat_full[!is.na(PadMat_full$symbol), ]
rownames(PadMat_full) <- PadMat_full$symbol

cat("Loaded expression matrix:", nrow(PadMat_full), "genes x", 
    ncol(PadMat_full) - 1, "samples\n\n")
```

---

## Section 4: Define Helper Functions

```{r helper-functions}
# ================================================================
# read_msigdb(): Parse GMT file into gene lists
# ================================================================
# Input: path to GMT file
# Output: List with two components:
#   $flat: Data frame with columns name, url, genes
#   $lists: Named list of gene vectors for each pathway
# ================================================================

read_msigdb <- function(path) {
  tmp <- sapply(readLines(path), function(x) {
    gsub(' ', '', strsplit(gsub('///', '\t', x), '\t')[[1]])
  })
  
  col <- vector(mode = 'character', length = length(tmp))
  out <- list(
    flat = data.frame(
      name = col, 
      url = col, 
      genes = col, 
      stringsAsFactors = FALSE
    ), 
    lists = list()
  )
  
  for (i in 1:length(tmp)) {
    out$flat[i, 'name'] <- tmp[[i]][1]
    out$flat[i, 'url'] <- tmp[[i]][2]
    out$flat[i, 'genes'] <- paste(tmp[[i]][3:length(tmp[[i]])], collapse = ", ")
    out$lists[i] <- list(tmp[[i]][3:length(tmp[[i]])])
  }
  
  names(out$lists) <- out$flat$name
  return(out)
}

# ================================================================
# munlist(): Flatten list of lists into indexed data frame
# ================================================================
# Helper function for working with nested list structures
# ================================================================

munlist <- function(x, force.rbind = FALSE, force.cbind = FALSE, force.list = FALSE) {
  if (!any(c(force.list, force.cbind, force.rbind))) {
    if (any(sapply(x, function(y) is.null(dim(y))))) {
      force.list <- TRUE
    }
    if (length(unique(sapply(x, function(y) dim(y)[2]))) == 1) {
      force.rbind <- TRUE
    }
    if ((length(unique(sapply(x, function(y) dim(y)[1]))) == 1)) {
      force.cbind <- TRUE
    }
  } else {
    force.list <- TRUE
  }
  
  if (force.list) {
    return(cbind(
      ix = unlist(lapply(1:length(x), function(y) rep(y, length(x[[y]])))),
      iix = unlist(lapply(1:length(x), function(y) if (length(x[[y]]) > 0) 1:length(x[[y]]) else NULL)),
      unlist(x)
    ))
  } else if (force.rbind) {
    return(cbind(
      ix = unlist(lapply(1:length(x), function(y) rep(y, nrow(x[[y]])))),
      iix = unlist(lapply(1:length(x), function(y) if (nrow(x[[y]]) > 0) 1:nrow(x[[y]]) else NULL)),
      do.call('rbind', x)
    ))
  } else if (force.cbind) {
    return(t(rbind(
      ix = unlist(lapply(1:length(x), function(y) rep(y, ncol(x[[y]])))),
      iix = unlist(lapply(1:length(x), function(y) if (ncol(x[[y]]) > 0) 1:ncol(x[[y]]) else NULL)),
      do.call('cbind', x)
    )))
  }
}
```

---

## Section 5: Run PADOG Analysis

This section loops through each gene set collection and each comparison, running PADOG and saving results.

```{r run-padog, message=TRUE, warning=FALSE}
# ================================================================
# MAIN ANALYSIS LOOP
# ================================================================

# Data frame to store all results across collections
total_padog_analysis <- data.frame()

# Loop through each gene set collection
for (gs_file in genesets_files) {
  
  # Extract collection name from filename
  # Example: "pathway_collection_v2023.gmt" → "pathway_collection"
  collection_name <- gs_file %>%
    basename() %>%
    str_remove("_gene_sets.gmt")
  
  cat("=" %in% collection_name, "\n")
  cat("Collection:", collection_name, "\n")
  
  # Create output subdirectory for this collection
  collection_output_dir <- file.path(output_dir, collection_name)
  dir.create(collection_output_dir, showWarnings = FALSE, recursive = TRUE)
  
  # Data frame to store results for current collection
  padog_by_collection <- data.frame()
  
  # ================================================================
  # Loop through each comparison
  # ================================================================
  
  for (comp in comparisons) {
    
    cat("  Comparison:", comp, "\n")
    
    # Extract disease and control groups from comparison string
    # "Bp_Ap" → disease="Bp", control="Ap"
    parts <- strsplit(comp, "_")[[1]]
    disease <- parts[1]
    control <- parts[2]
    
    # Select samples belonging to either group
    samples_sel <- metadata$Sample[metadata$gp %in% c(control, disease)]
    
    # Order samples: control first, then disease (for design matrix)
    samples_ordered <- samples_sel[
      order(factor(metadata$gp[samples_sel], levels = c(control, disease)))
    ]
    
    # Extract submatrix for selected samples
    PadMat_subset <- as.matrix(PadMat_full[, samples_ordered, drop = FALSE])
    
    # Remove genes with no counts
    PadMat_subset <- PadMat_subset[rowSums(PadMat_subset) > 0, ]
    
    # Load gene sets for this collection
    msigdb <- read_msigdb(path = gs_file)
    gslist <- msigdb$lists
    
    # Create group vector: "c"=control, "d"=disease
    group_vector <- ifelse(metadata[samples_ordered, "gp"] == control, "c", "d")
    
    # Create design matrix for voom normalization
    design <- model.matrix(~ group_vector)
    
    # Voom normalization (transforms count data to log-scale)
    y <- voom(PadMat_subset, design, plot = FALSE)
    
    # Convert group_vector to factor
    group_vector <- factor(group_vector, levels = c("c", "d"))
    
    # ================================================================
    # Run PADOG
    # ================================================================
    # Parameters:
    #   Nmin: Minimum number of genes in a pathway (default: 3)
    #   NI: Number of permutations (default: 500)
    #   ncr: Number of cores for parallel processing
    
    padog_analysis <- padog(
      esetm = y$E,
      group = group_vector,
      gslist = gslist,
      organism = "hsa",
      verbose = TRUE,
      Nmin = 3,           # Minimum genes per pathway
      NI = 500,           # Number of permutations
      plots = FALSE,
      dseed = 3,
      parallel = TRUE,
      ncr = 2             # Number of cores (adjust based on your system)
    )
    
    # Adjust p-values for multiple testing (FDR)
    padog_analysis$Adj <- p.adjust(padog_analysis$Ppadog, method = "fdr")
    
    # Add metadata columns
    padog_analysis$Collection <- collection_name
    padog_analysis$Comparison <- comp
    
    # Combine results
    padog_by_collection <- rbind(padog_by_collection, padog_analysis)
  }
  
  # ================================================================
  # Save results for this collection
  # ================================================================
  
  output_file <- file.path(
    collection_output_dir, 
    paste0(collection_name, ".csv")
  )
  write.csv(
    padog_by_collection, 
    output_file, 
    quote = FALSE, 
    row.names = FALSE
  )
  
  cat("    → Saved:", output_file, "\n\n")
  
  # Add to total results
  total_padog_analysis <- rbind(total_padog_analysis, padog_by_collection)
}

# ================================================================
# Save Total Results
# ================================================================

write.csv(
  total_padog_analysis,
  file.path(output_dir, "total_padog_analysis.csv"),
  quote = FALSE,
  row.names = FALSE
)

saveRDS(
  total_padog_analysis,
  file.path(output_dir, "total_padog_analysis.rds")
)

cat("Total results saved:\n")
cat("  CSV: ", file.path(output_dir, "total_padog_analysis.csv"), "\n")
cat("  RDS: ", file.path(output_dir, "total_padog_analysis.rds"), "\n")
```

---

## Section 6: Filter Significant Results

```{r filter-significant}
# ================================================================
# Filter pathways with adjusted p-value < 0.25 in Bp_Ap comparison
# ================================================================

sig_bp_ap_genes <- total_padog_analysis %>%
  filter(Comparison == "Bp_Ap", Adj < 0.25) %>%
  pull(Name) %>%
  unique()

cat("Significant pathways (FDR < 0.25) in Bp_Ap comparison:", 
    length(sig_bp_ap_genes), "\n\n")

# Create filtered dataset with significance annotations
sig_sets <- total_padog_analysis %>%
  filter(Name %in% sig_bp_ap_genes) %>%
  select(-ID) %>%
  mutate(significance = ifelse(Adj < 0.25, "sig", "not sig"))

# Save filtered results
write.csv(
  sig_sets,
  file = file.path(output_dir, "PADOG_FDR_0.25_in_Bp_Ap.csv"),
  row.names = FALSE
)

saveRDS(
  sig_sets,
  file = file.path(output_dir, "PADOG_FDR_0.25_in_Bp_Ap.rds")
)

cat("Filtered results saved.\n\n")

# Summary statistics
cat("Summary of significant pathways:\n")
cat("  Total pathways:", length(unique(sig_sets$GeneSet)), "\n")
cat("  By direction:\n")
print(table(sig_sets[sig_sets$Comparison == "Bp_Ap", "Direction"]))
```

---

## Notes

- **Parallelization:** Adjust `ncr` parameter in PADOG call to match your CPU cores
- **Thresholds:** FDR cutoff (0.25) can be adjusted based on your stringency requirements
- **Comparisons:** Ensure comparison identifiers match exactly with values in metadata$gp
- **Missing values:** Genes without symbols and zero-count genes are automatically removed

---

## Session Info

```{r session-info}
sessionInfo()
```
